# (PART) Introduction to R {-}

# Language essentials {#chapter1}

## Why R?

R is an open-source statistical programming language used worldwide for statistical analysis and data science in a vast variety of fields. The R project was released to the public in 2000 and since then *The R Foundation* and the *R Development Core Team* take care of development of the language and of the release of new versions.^[For an interesting but a bit technical presentation of the R language see the article "R: A language for data analysis and graphics" written by two of R's main developers, Ross Ihaka and Robert Gentleman [@Ihaka].] Unlike many other computer programs for statistical and data analysis, in R you don't perform your operations by clicking. In R you work with a script, basically a text where the user write functions to perform the desired operations, such as computing averages, transforming data and so on. Although at first this seems very difficult especially for those who do not have experience with code based work, writing your own operations as if it were normal text has many advantages. This allows you not only to think carefully about the operations you are performing (rather than just clicking left and right) but also to ensure the reproducibility of your work as the academic and scientific work demand. Another very good reason for using R is the large availability of R packages. R packages are collections of functions, data and other code that extend the capabilities of "base" R. Nowadays there are many R packages available to perform a wide variety of tasks, from data visualization, simple and sophisticated econometric analysis, web application development and much more. Not least, R's ability to visualize data with publication quality graphs and to write data-driven texts make it an ideal choice for economists. RStudio

## Basic operations

R can be used as a calculator

```{r}
2 + 2
```

```{r}
exp(0)
```

```{r}
log(1)
```

Assigning a value to a variable (short cut Alt -). The object `x` is saved in the environment.

```{r}
x <- 15
```

To remove a variable

```{r}
rm(x)
```

R handle data vectors as single objects. We can easily create data vectors

```{r}
x <- c(1, 2, 3)
```

We can see the object names that we have already created

```{r}
ls()
```

We can easily remove all objects from the current workspace
It is a good practice to start every new R Session with this line of code

```{r}
rm(list = ls())
```

R can easily creates sequences of numbers

```{r}
a <- 1:10 
b <- seq(from = 1, to = 10, by = 1)
```

Generating random numbers from a normal distribution specifying mean and standard deviation. The `set.seed` function allows us to generate the same numbers. This function is essential if we want someone else to be able to replicate the results of our simulations, for example when we want to share our work with colleagues or our  thesis supervisor or when for example when we want to publish a paper.

```{r}
set.seed(1234)

rnorm(n = 10, mean = 0, sd = 1)
```

Assigning (a vector of) random numbers to a variable called `y`. The object `y` is now saved in the environment. 

```{r}
y <- rnorm(10, 0, 1)
```

A very important function in R that we must absolutely know is the `length` function. The `length` function tells us the number of elements in a vector (but also other types of objects). We apply the `length` function to the `y` object that we created a moment ago. The result is obviously `10`.

```{r}
length(y)
```

Logical Operations. The operator "==" force R to make a "true or false" judgement. Be careful because the sign `=` in R can be used to define a variable, e.g. `x = 5`. This is why I recommend using the assignment operator `<-` to define variables and use the equal sign where you cannot do otherwise.

```{r}
2 + 2 == 4
sqrt(25) == 4
```

R also includes the normal arithmetic and logical operators.

```{r}
100 >= 100
99 > 100
99 != 100
!(1==1) # One is not equal to one -> FALSE!
(1==1) | (2==3) # One is equal to one OR two is equal to three -> TRUE!
(1==1) & (2==3) # One is equal to one AND two is equal to three -> FALSE!
```

- basic functions (double equal, root, logical operators)

## 

## Basic data structures

Vector, Matrix, Dataframe, List

## For loops

```{r}

```

## Nested loops

```{r}

data_interests <- matrix(nrow = 100, ncol = 3, byrow = TRUE)

interests <- c(0.01, 0.02, 0.03) 

data_interests[1,1] <- 1
data_interests[1,2] <- 1
data_interests[1,3] <- 1

for (t in 2:nrow(data_interests)) {
  
  for (s in 1:ncol(data_interests)) {
    
    data_interests[t,s] <- data_interests[t-1,s] * (1 + interests[s])
    
  }
  
}

```

```{r}

plot(NULL, 
     xlim = c(1, nrow(data_interests)), 
     ylim = c(0, max(data_interests)),
     main = "Investments", 
     xlab = "Year", 
     ylab = "Euro")

for (l in 1:ncol(data_interests)) {
lines(data_interests[,l], type = "l", col = l)
}
```

## Simple functions

## The apply family

The apply "family" of functions.


We have already reached the end of the first chapter. I hope you are not bored. In the next chapter, we will take a closer look at how to work with data in a dataframe, such as how to edit data, select variables, create new ones and much more.






























