# (PART) Introduction to R {-}

# Code 1 {#chapter1}

## Loading and managing data

To begin with, let's see how to install a new package in R. We are going to install the *wooldridge* package which contains all the datasets used in the book *Introductory Econometrics* by Jeffrey M. Wooldridge. Additional information regarding the datasets can be found in the following [document](https://cran.r-project.org/web/packages/wooldridge/wooldridge.pdf). 

```{r, eval = FALSE}
# Install the package
install.packages("wooldridge")

# Upload the package to the library
library(wooldridge)
```

Now, let us see how to unload a package from the library. The quickest way is to untick the checkbox of the package you want to remove in the *Packages* window. The following code is automatically executed in the console. There is no need to copy it or run it again. Another, perhaps more radical, way to unload all packages is to restart the RStudio session. In this case, the keyboard shortcut *CTRL + SHIFT + F10* can be used.

```{r, eval = FALSE}
# Unloading a package
detach("package:wooldridge", unload = TRUE)
```

After loading the dataset, in our case contained in the *wooldridge* package, we can take a look at the data. It is important to take a look at the data, for example after having modified existing variables or having created new ones. The dataset we use in this exercise is *wage1*. We can take a glance at the data using the *View* function.

```{r, eval = FALSE}

View(wage1)

data("wage1")

# If you have your data stored as CSV, we can import the dataset using the read_csv function from the readr package. This is the "evolution" of the function read.csv
# Remember to use the forward slash / or double backslash \\ in R
library(readr)

wage1 <- read_csv("wage1.csv")

# Alternatively, you can use the import function in the Environment which produces this code:

```

## Data manipulation using base R

```{r, eval = FALSE}

# Indexing when you need to address a particular element into a vector, for example the wage of worker nr. 5
educ[5]

# If we want data for more than one worker
educ[c(2,3,5)]

# Data from worker one to worker five
educ[1:5]

# If we want to modify one particular observation 
educ[5] <- NA

# Negative indexing allow to show all data except those specified in parenthesis
educ[-c(1:520)]

# Negative indexing allow us to drop specified rows and columns
# Drop the first row in the dataset
wage1[-1,]

# Drop the last column in the dataset
wage1[,-25]

# Just check the number of row and col in wage1 with the dim() function
dim(wage1)

# Converting dollars to euros and add the new column variable to the dataset
wage1$wage_EUR <- wage * 0.86

```

## Conditional selection

```{r, eval = FALSE}

# To use when you need to extract some data that satisfy certain criteria
# Workers that have more than 15 years of education
educ[educ > 15]

# Workers that have between 15 AND 18 year of education
educ[educ >= 15 & educ <= 18]

# Workers that have between 15 OR 18 year of education
educ[educ >= 17 | educ <= 2]

# A more interesting research question is to ask the hourly wage of workers that have 15 or more years of education
wage[educ >= 15]

# Hourly wage of workers with more than 15 years of education and 5 years of experience 
wage[educ >= 15 & exper >= 5]

# Look at what happens when we type educ>15. We can exploit the fact that TRUE=1 and FALSE=0 to calculate
# the percentage of workers that have more than 15 years of education
sum(educ > 15) / sum(educ >= 0) * 100

```

